---
title: "BRISC extensions checks"
author: "Lukas Weber"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 3
    toc_float:
      collapsed: false
---

```{r setup, eval=TRUE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


# Background

This document runs several short examples to check that our updates in the `BRISC-extensions` repository (return log likelihoods, re-use ordering of coordinates, re-use nearest neighbors) do not break any existing results in unexpected ways.

```{r, message=FALSE}
library(BRISC)
```


# Example 1: BRISC_estimation documentation

Example from `BRISC_estimation()` function documentation.


## Original example

```{r}
rmvn <- function(n, mu = 0, V = matrix(1)){
  p <- length(mu)
  if(any(is.na(match(dim(V),p))))
    stop("Dimension not right!")
  D <- chol(V)
  t(matrix(rnorm(n*p), ncol=p)%*%D + rep(mu,rep(n,p)))
}

set.seed(1)
n <- 1000
coords <- cbind(runif(n,0,1), runif(n,0,1))

beta <- c(1,5)
x <- cbind(rnorm(n), rnorm(n))

sigma.sq = 1
phi = 1
tau.sq = 0.1

B <- as.matrix(beta)
D <- as.matrix(dist(coords))
R <- exp(-phi*D)
w <- rmvn(1, rep(0,n), sigma.sq*R)

y <- rnorm(n, x%*%B + w, sqrt(tau.sq))

estimation_result <- BRISC_estimation(coords, y, x)
estimation_result$Theta ##Estimates of covariance model parameters.
estimation_result$Beta ##Estimates of Beta
```


## Re-use ordering

```{r}
# calculate ordering
order_brisc <- BRISC_order(coords)

# calculate results with ordering argument
estimation_result_order <- BRISC_estimation(coords, y, x, 
                                            ordering = order_brisc)
estimation_result_order$Theta
estimation_result_order$Beta

# check if results equal
all(estimation_result_order$Theta == estimation_result$Theta)
all(estimation_result_order$Beta == estimation_result$Beta)
```


## Re-use neighbors

```{r}
# calculate neighbors
nn_brisc <- BRISC_neighbor(coords)

# calculate results with neighbors argument
estimation_result_nn <- BRISC_estimation(coords, y, x, 
                                         neighbor = nn_brisc)
estimation_result_nn$Theta
estimation_result_nn$Beta

# check if results equal
all(estimation_result_nn$Theta == estimation_result$Theta)
all(estimation_result_nn$Beta == estimation_result$Beta)
```


## Re-use ordering and neighbors

```{r}
# calculate neighbors with ordering argument
nn_brisc_order <- BRISC_neighbor(coords, ordering = order_brisc)

# check if neighbors object identical with and without ordering argument
identical(nn_brisc, nn_brisc_order)

# calculate results with ordering and neighbors arguments
estimation_result_order_nn <- BRISC_estimation(coords, y, x, 
                                               ordering = order_brisc, 
                                               neighbor = nn_brisc)
estimation_result_order_nn$Theta
estimation_result_order_nn$Beta

# check if results equal
all(estimation_result_order_nn$Theta == estimation_result$Theta)
all(estimation_result_order_nn$Beta == estimation_result$Beta)
```


# Example 2: Human DLPFC

Example using spatial transcriptomics data from human dorsolateral prefrontal cortex (DLPFC) dataset (Maynard and Collado-Torres et al. 2021).


## nnSVG: re-using ordering and neighbors

Using code from nnSVG wrapper to run BRISC on a subset of 21 genes in human DLPFC dataset.

By default, the nnSVG wrapper re-uses both ordering and neighbors for faster runtime.

Running code manually (instead of using wrapper) so we can put `set.seed()` in the same places.

```{r, message=FALSE}
library(SpatialExperiment)
library(STexampleData)
library(nnSVG)
library(BRISC)
library(BiocParallel)
```

```{r}
# load example dataset
spe <- Visium_humanDLPFC()
dim(spe)

# preprocessing steps
# set seed for reproducibility
set.seed(123)
spe <- preprocessSVG(spe)
dim(spe)

# subsampling to select small number of random genes and one known SVG for 
# faster runtime in example
set.seed(123)
ix_20 <- sample(seq_len(nrow(spe)), 20)
ix_snap25 <- which(rowData(spe)$gene_name == "SNAP25")
ix <- c(ix_20, ix_snap25)
spe <- spe[ix, ]
dim(spe)
```

```{r}
# run BRISC via nnSVG code
# adapting code from nnSVG wrapper at https://github.com/lmweber/nnSVG/blob/main/R/nnSVG.R

# ---------
# run BRISC
# ---------

y <- assays(spe)[["binomial_deviance_residuals"]]

# scale coordinates proportionally
coords <- spatialCoords(spe)
range_all <- max(apply(coords, 2, function(col) diff(range(col))))
coords <- apply(coords, 2, function(col) (col - min(col)) / range_all)

# calculate ordering of coordinates
order_brisc <- BRISC_order(coords, order = "AMMD")

# calculate nearest neighbors
nn_brisc <- BRISC_neighbor(coords, n.neighbors = 15, n_omp = 1, 
                           search.type = "cb", ordering = order_brisc)

# run BRISC using parallelization
ix <- seq_len(nrow(y))
out_brisc <- bplapply(ix, function(i) {
  # fit model (intercept-only model if x is NULL)
  y_i <- y[i, ]
  suppressWarnings({
    runtime <- system.time({
      # set seed for reproducibility
      set.seed(123)
      out_i <- BRISC_estimation(coords = coords, y = y_i, x = NULL, 
                                cov.model = "exponential", 
                                ordering = order_brisc, neighbor = nn_brisc, 
                                verbose = FALSE)
    })
  })
  res_i <- c(
    out_i$Theta, 
    loglik = out_i$log_likelihood, 
    runtime = runtime[["elapsed"]]
  )
  res_i
}, BPPARAM = MulticoreParam(workers = 4))

# collapse output list into matrix
mat_brisc <- do.call("rbind", out_brisc)
```


## nnSVG: without re-using ordering and neighbors

Run code from nnSVG wrapper without re-using ordering and neighbors.

```{r}
# run BRISC via nnSVG code
# adapting code from nnSVG wrapper at https://github.com/lmweber/nnSVG/blob/main/R/nnSVG.R

# ---------
# run BRISC
# ---------

y <- assays(spe)[["binomial_deviance_residuals"]]

# scale coordinates proportionally
coords <- spatialCoords(spe)
range_all <- max(apply(coords, 2, function(col) diff(range(col))))
coords <- apply(coords, 2, function(col) (col - min(col)) / range_all)

# run BRISC using parallelization
ix <- seq_len(nrow(y))
out_brisc_no_order_nn <- bplapply(ix, function(i) {
  # fit model (intercept-only model if x is NULL)
  y_i <- y[i, ]
  suppressWarnings({
    runtime <- system.time({
      # set seed for reproducibility
      set.seed(123)
      out_i <- BRISC_estimation(coords = coords, y = y_i, x = NULL, 
                                cov.model = "exponential", 
                                order = "AMMD", 
                                n.neighbors = 15, search.type = "cb", 
                                verbose = FALSE)
    })
  })
  res_i <- c(
    out_i$Theta, 
    loglik = out_i$log_likelihood, 
    runtime = runtime[["elapsed"]]
  )
  res_i
}, BPPARAM = MulticoreParam(workers = 4))

# collapse output list into matrix
mat_brisc_no_order_nn <- do.call("rbind", out_brisc)
```


## Compare results

```{r}
# default: re-use ordering and neighbors
mat_brisc

# without re-using ordering and neighbors
mat_brisc_no_order_nn

# check if identical
identical(mat_brisc, mat_brisc_no_order_nn)
```


# Session info

```{r}
sessionInfo()
```

