---
title: "BRISC extensions checks (base)"
author: "Lukas Weber"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 3
    toc_float:
      collapsed: false
---

```{r setup, eval=TRUE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


# Background

This document runs several short examples to check that our updates in the `BRISC-extensions` repository (return log likelihoods, re-use ordering of coordinates, re-use nearest neighbors) do not break any existing results in unexpected ways.

```{r, message=FALSE}
library(BRISC)
```


## Example 1: BRISC_estimation documentation

Example from `BRISC_estimation()` function documentation.

Run using current CRAN version of BRISC (without new arguments for ordering and neighbors) to compare results against new version.


### Original example

```{r}
rmvn <- function(n, mu = 0, V = matrix(1)){
  p <- length(mu)
  if(any(is.na(match(dim(V),p))))
    stop("Dimension not right!")
  D <- chol(V)
  t(matrix(rnorm(n*p), ncol=p)%*%D + rep(mu,rep(n,p)))
}

set.seed(1)
n <- 1000
coords <- cbind(runif(n,0,1), runif(n,0,1))

beta <- c(1,5)
x <- cbind(rnorm(n), rnorm(n))

sigma.sq = 1
phi = 1
tau.sq = 0.1

B <- as.matrix(beta)
D <- as.matrix(dist(coords))
R <- exp(-phi*D)
w <- rmvn(1, rep(0,n), sigma.sq*R)

y <- rnorm(n, x%*%B + w, sqrt(tau.sq))

estimation_result <- BRISC_estimation(coords, y, x)
estimation_result$Theta ##Estimates of covariance model parameters.
estimation_result$Beta ##Estimates of Beta
```


# Session info

```{r}
sessionInfo()
```

